---
title: "PS 3 Week 3 Section"
subtitle: "Subsetting, Means, and One-/Two-Way Tables"
author: "PS 3 Section"
date: "Week 3"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "seahorse"
    fonttheme: "structurebold"
    slide_level: 2
params:
  show_answers: true
header-includes:
  - \usepackage{array}
  - \definecolor{darkgreen}{rgb}{0, 0.5, 0}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = "#>", size = "small")

# Change this filename each week to that week's QR code image.
# Just drop the image file in the same folder as this .Rmd.
qr_image <- "attendance_qr.png"
```

## Today's Agenda

- Vocabulary & key terms recap
- Functions
- Live demo: putting it together
- Common mistakes (and how to avoid them)
- Why do we care? Tables, hypotheses, and a caution

*Goal: by the end of section, you can go from a raw dataset to an answer
to a research question, using `$`, `mean()`, `subset()`, and `table()`.*

## Quick Vocabulary Check

\small

- **Data frame:** a rectangular dataset in R -- rows are observations,
  columns are variables (e.g. `cafe`).
- **Variable (column):** one measured characteristic, e.g. `drinks`.
- **Observation (row):** one unit in the dataset, e.g. one day/time
  slot.
- **Scalar:** a single value, e.g. `85`, or the result of `mean(x)`.
- **Vector:** a sequence of values of the same type, e.g.
  `cafe$drinks`.
- **Categorical variable:** takes on named categories, e.g. `"AM"` /
  `"PM"`.
- **Function / argument:** `mean(x)` is a function; `x` is the
  argument you give it.


## Functions we have learned

\footnotesize
\begin{center}
\begin{tabular}{|p{2cm}|p{3.4cm}|p{4.8cm}|}
\hline
\textbf{Function} & \textbf{What it does} & \textbf{Example} \\
\hline
\texttt{dataset\$var} & pulls out one column & \texttt{officerdata\$contra} \\
\hline
\texttt{mean(x)} & average of a numeric variable & \texttt{mean(officerdata\$contra)} \\
\hline
\texttt{subset(data, cond)} & keeps only rows meeting a condition & \texttt{subset(officerdata, officer\_female == 1)} \\
\hline
\texttt{table(x)} & counts of one categorical variable & \texttt{table(officerdata\$driver\_race)} \\
\hline
\texttt{table(x, y)} & counts of combinations of two categorical variables & \texttt{table(x, y)} \\
\hline
\end{tabular}
\end{center}

## A New Example: Coffee Shop Data

Small dataset practice.

```{r cafe-data, size="footnotesize"}
cafe <- data.frame(
  day      = c("Mon","Mon","Tue","Tue","Wed","Wed","Fri","Fri"),
  daypart  = c("AM","PM","AM","PM","AM","PM","AM","PM"),
  drinks   = c(85,40,90,45,78,50,120,80),
  size_cat = c("Rush","Slow","Rush","Slow","Rush","Slow","Rush","Rush")
)
head(cafe)
```

`daypart`: `"AM"` = morning, `"PM"` = afternoon.

## Spot the error (or lack thereof)

Discuss these in small groups (2-4 people) and if you think it will run and if it does, what will it print out.

```{r error-1, eval=FALSE}
cafe.morning <- subset(cafe, daypart == "AM")
```


```{r error-2, eval=FALSE}
cafe$cafe.morning$drinks
```


```{r error-3, eval=FALSE}
mean(cafe)
```

## Spot the error (or lack thereof)

```{r error-1-eval, eval=FALSE}
cafe.morning <- subset(cafe, daypart == "AM")
```

\textcolor{darkgreen}{Correct!}

```{r error-2-eval, eval=FALSE}
cafe$cafe.morning$drinks
```

\textcolor{red}{Should just be }`cafe.morning$drinks`

```{r error-3-eval, eval=FALSE}
mean(cafe)
```

\textcolor{red}{Needs to do a specific column in the dataset. Such as }`mean(cafe$drinks)`


## Step 1: Pull Out a Variable and Take a Mean

```{r mean-demo}
cafe$drinks
mean(cafe$drinks)
```

**What does this one number tell us and what does it hide?**

## Step 2: Subset, Then Summarize

```{r subset-demo}
morning.only <- subset(cafe, daypart == "AM")
mean(morning.only$drinks)
```

- Same idea as Week 2: `subset()` first, *then* compute on the subset.
- What changes when we switch `"AM"` for `"PM"`?

## What changes when we switch "AM" for "PM"?

```{r subset-demo-afternoon}
afternoon.only <- subset(cafe, daypart == "PM")
mean(afternoon.only$drinks)
```

## Step 3: One-Way Table

```{r oneway-demo}
table(cafe$size_cat)
```

- Frequency table: how many rows fall into each category?
- Useful first step before you subset or cross-tabulate.

## Step 4: Two-Way Table

```{r twoway-demo}
table(cafe$daypart, cafe$size_cat)
```

- Rows = first variable, columns = second variable.
- How many instances fall into combinations of categories?

## Correcting Common Mistakes

```{r mistakes-demo, eval=FALSE}
# Mistake: single = instead of double ==
subset(cafe, daypart = "AM")      # ERROR

# Fixed: use == to test equality, not =
subset(cafe, daypart == "AM")     # correct
```

- `=` **assigns** a value; `==` **compares** two values.
- Only `==` belongs inside a `subset()` condition.

## Mean of a Categorical Variable

```{r categorical-mean-demo, warning=TRUE}
mean(cafe$daypart)
```

- `daypart` is categorical (`"AM"` / `"PM"`), not numeric. So `mean()` gives a warning and returns `NA`
- Check whether a variable is categorical or numeric *before* you use `mean()`

## Using the $ to get a column from a dataframe

- Before the `$` is the name of the dataframe, after the `$` is the name of the column.
- If we wanted to get the days in the dataset, what would we type?
   - column names are: `day`, `daypart`, `drinks`, `size_cat`
   - name of the dataframe: `cafe`

## Using the $ to get a column from a dataframe
- If we wanted to get the days in the dataset, what would we type?

```{r column-subset, warning=TRUE}
cafe$day
```

## Why Do We Care? (Recap from Lecture)

$$\text{Theory} \rightarrow \text{Hypothesis} \rightarrow \text{Empirical Test} \rightarrow \text{Re-Evaluate Theory}$$

- A one-way table describes *one* thing.
- A two-way table lets us **test a hypothesis** about how two variables relate.
- Example: "Rush periods are more common in the morning" is a hypothesis we can *check* with the two-way table above.

## Tips for Analyzing the Data

- Stick to the facts first: what does the table literally say?
- *Then* you might hypothesize about *why*.
- **We only know what the data shows.** The data can't tell us why something is happening.
  - E.g., with the officer-stop data from Week 2: if female officers
    have a higher contraband-find rate than male officers, is that
    because they're better at spotting contraband, or because
    they're more selective about who they search? The table alone
    can't say.
    
## Reminders

- Make sure to follow the instructions at the end of the assignments before you submit
   - If you forget to save and wait 5 seconds, you may lose points because it does not have the last things you typed in
- Make sure to use the same names of dataframes and variables as the original .ipynb file has -- that is how the autograder is checking your work

## Recap

\footnotesize
\begin{center}
\begin{tabular}{|p{2.8cm}|p{6.6cm}|}
\hline
\textbf{Function} & \textbf{Use it when...} \\
\hline
\texttt{dataset\$var} & you need one column \\
\hline
\texttt{mean()} & you want the average of a numeric variable \\
\hline
\texttt{subset()} & you want only some rows \\
\hline
\texttt{table(x)} & you want counts of one categorical variable \\
\hline
\texttt{table(x, y)} & you want to see how two categorical variables relate \\
\hline
\end{tabular}
\end{center}

## Before You Go

- You have the remainder of class to work on the problem set.
\vspace{1em}
- **Problem Set 1:** due **October 2**.
- R Cheat Sheet has every function from today.
- My office hours: **Mon & Fri, 1--2pm**.
- Check the syllabus for other GSIs' office hours.

## Attendance

\begin{center}
\includegraphics[width=0.5\textwidth]{`r qr_image`}
\end{center}
