---
title: "PS 3: Week 5"
subtitle: "Randomized Experiments vs. Observational Data"
author: "Maya Mikelson"
date: "Week 5"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "default"
    slide_level: 2
params:
  show_answers: true
header-includes:
  - \usepackage{array}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = "#>", size = "small")

social <- read.csv('ps3_week5_social_pressure.csv')

# Change this filename each week to that week's QR code image.
# Just drop the image file in the same folder as this .Rmd.
qr_image <- "attendance_qr.png"
```

## A Motivating Case: Tylenol \& Autism

- Some observational data shows children whose mothers took
  acetaminophen during pregnancy were slightly more likely to be
  diagnosed with autism.
- \textcolor{blue}{Is this causal?}

\pause

- **Omitted variable bias:** people who take Tylenol may do so
  because of other conditions (chronic pain, environmental factors,
  etc.) that could *also* raise autism risk.
- **Selection bias:** people who choose to take a "treatment" may
  differ, in ways connected to the outcome, from those who don't.

## A Potential Causal Effect: Populism

- Campaigns want to know whether **populist rhetoric** -- framing
  some "elite" (a party, billionaires, cultural elites) as unfairly
  ruling over "the people" -- actually persuades voters.
\pause
- Imagine voters in a 2026 campaign: some are exposed to populist
  rhetoric, and each later votes for Party P or Party O.

\pause

\vspace{6pt}
**Discuss:**

- \textcolor{blue}{What's the \textbf{treatment} in this example?}
\vspace{.5em}
- \textcolor{blue}{What are the \textbf{potential outcomes}?}
\vspace{.5em}
- \textcolor{blue}{What is each individual's \textbf{treatment effect}?}

```{r populism, results='asis', echo=FALSE}
if (params$show_answers) {
  cat("
## Populism

- \\textcolor{blue}{What's the \\textbf{treatment} in this example?}

\\begin{quote}
Exposure to populist rhetoric
\\end{quote}

- \\textcolor{blue}{What are the \\textbf{potential outcomes}?}

\\begin{quote}
Voters vote choice when exposed to populist rhetoric

Voters vote choice when not exposed to populist rhetoric
\\end{quote}

- \\textcolor{blue}{What is each individual's \\textbf{treatment effect}?}

\\begin{quote}
Vote when exposed to populist rhetoric - Vote when not exposed to populist rhetoric
\\end{quote}
")
}
```


## Potential Outcomes

\footnotesize
\begin{center}
\begin{tabular}{|p{1.6cm}|p{2.5cm}|p{3cm}|p{3cm}|}
\hline
\textbf{Voter} & \textbf{Vote if exposed?} & \textbf{Vote if not exposed?} & \textbf{Individual Treatment Effect} \\
\hline
Tess & 0 & 0 & 0 \\
\hline
Jeremy & 1 & 1 & 0 \\
\hline
Anne Marie & 1 & 0 & 1 \\
\hline
Jack & 0 & 1 & -1 \\
\hline
\end{tabular}
\end{center}

*1: Party P*

*0: Party O*

## Observational Data on Populism

- A survey of over 1 million voters asks: were you exposed to Party
  P's populist rhetoric in 2026 (TV, social media, mailers, etc.)?
  Who did you vote for?
- People exposed to the rhetoric are **far more likely** to vote for
  Party P.

\pause

\vspace{6pt}
**Discuss:**

- \textcolor{blue}{How confident should we be that this is \textit{causal}?}
- \textcolor{blue}{What sources of bias might undermine this finding?}
- \textcolor{blue}{What are the \textit{advantages} of a dataset like this?}


```{r obs-populism, results='asis', echo=FALSE}
if (params$show_answers) {
  cat("
## Observational Data on Populism

- \\textcolor{blue}{How confident should we be that this is \\textit{causal}?}

\\begin{quote}
\\small Not confident. There is no random assignment. People could have opted in to mailers.
\\end{quote}

- \\textcolor{blue}{What sources of bias might undermine this finding?}

\\begin{quote}
\\small \\textbf{Selection Bias}: People who remembered receiving populist rhetoric might have been more influenced by it and therefore also more likely to report voting for Party P.

\\small \\textbf{Omitted Variable Bias}: Underlying factors (like prior political leaning, geographic location, or media consumption) could cause both exposure to the rhetoric and the decision to vote for Party P.

\\small \\textbf{Reverse Causation}: Voters who already decided to vote for Party P might actively seek out their campaign materials or sympathetic media.
\\end{quote}

- \\textcolor{blue}{What are the \\textit{advantages} of a dataset like this?}

\\begin{quote}
\\small \\textbf{Massive sample size}: 1 million voters provides high statistical power.

\\small \\textbf{High external validity}: It captures how real voters behave in the real world during an actual campaign, rather than in an artificial lab setting.
\\end{quote}

")
}
```


## Omitted Variable Bias, Selection Bias, and Reverse Causation!

\footnotesize
\begin{center}
\begin{tabular}{|p{1.4cm}|p{2.8cm}|p{2.6cm}|p{2.6cm}|}
\hline
\textbf{Voter} & \textbf{Exposed in 2025?} & \textbf{Voted P 2025?} & \textbf{Voted P 2021?} \\
\hline
Tess & 1 & 1 & 1 \\
\hline
Jeremy & 1 & 1 & 1 \\
\hline
Anne Marie & 0 & 0 & 0 \\
\hline
Jack & 0 & 0 & 0 \\
\hline
\end{tabular}
\end{center}

## How Do We Resolve These Biases?

- With observational data, people can **select into treatment**.
- If that selection *and* the outcome are both explained by some
  other variable, that's **omitted variable bias**.
  - E.g., men being less risk-averse $\rightarrow$ more likely to
    take an experimental drug *and* have lower resting heart rates.
- **Randomization** makes treatment and control groups similar on
  *all* variables (given a large enough sample) -- no more omitted variable bias to
  worry about.
  - E.g., prior vote can no longer cause *both* exposure to rhetoric
    and future vote, because it's balanced across groups.

## Designing an Experiment on Populism

**How might we set this up?**

- What's the treatment?
- How is the treatment assigned?
- What outcome(s) are we measuring?
- What variables should we measure?
  - Often driven by which **omitted variables** we're worried about.
  
  
```{r populism-experiment, results='asis', echo=FALSE}
if (params$show_answers) {
  cat("
## Designing an Experiment on Populism

- \\textcolor{blue}{What's the treatment?}

\\begin{quote}
\\textbf{Treatment group} sees a message from Party P: Party O has spent 12 years in power doing nothing for hardworking people while enriching themselves and their billionaire friends; Party P is committed to social welfare spending instead.
  
\\textbf{Control group} sees nothing.
\\end{quote}

- \\textcolor{blue}{How is the treatment assigned?}

\\begin{quote}
Randomly
\\end{quote}

- \\textcolor{blue}{What outcome(s) are we measuring?}

\\begin{quote}
Who they'd vote for in a hypothetical P vs. O election
\\end{quote}

- \\textcolor{blue}{What variables should we measure?}

\\begin{quote}
Past vote history, demographics, political knowledge, and political attention
\\end{quote}

")
}
```


## Bundled Treatments

- To find a causal effect, you need to **isolate** your explanatory
  variable from your outcome.
- If your treatment actually bundles *multiple* things together, how
  do you know which part is doing the work?
  - Is it the **populism**? Or the **policies**?
- Solution: split the bundle across multiple treatment arms.
  - Populism treatment (previous slide)
  - Policies only, no populist framing
  - Control group
- Compare the bundled treatment to the "policies only" arm to
  isolate the effect of populist rhetoric *itself*.


```{r attendance, results='asis', echo=FALSE}
if (params$show_answers) {
  cat(paste0("
## Attendance Survey

\\begin{center}
\\includegraphics[width=0.45\\textwidth]{", qr_image, "}
\\end{center}
"))
}
```

## Practice Coding Exercise 1

- Note: \textit{This is a harder exercise than we have done in class or in the problem set before.}

\pause

- Using the data set on persuasion and voting from Wednesday in `Week5_Sandbox1.ipynb`



- What is the turnout percentage for each treatment? (Civic Duty, Hawthorne, Self, and Neighbors)



## Answer: Coding Exercise 1 {.fragile}

\footnotesize
```{r answer-data, echo=TRUE, eval=params$show_answers, include=params$show_answers}
civic.data <- subset(social, treat_civic == 1)
hawthorne.data <- subset(social, treat_hawthorne == 1)
self.data <- subset(social, treat_self == 1)
neighbors.data <- subset(social, treat_neighbors == 1)
```

```{r answer-q1-method1, echo=TRUE, eval=params$show_answers, include=params$show_answers}
civic.turnout <- mean(civic.data$outcome_voted)
hawthorne.turnout <- mean(hawthorne.data$outcome_voted)
self.turnout <- mean(self.data$outcome_voted)
neighbors.turnout <- mean(neighbors.data$outcome_voted)

c(civic.turnout, hawthorne.turnout, self.turnout, neighbors.turnout)
```

## Answer: Coding Exercise 1 {.fragile}

We could also do it all in one step:

\footnotesize
```{r answer-q1-method2, echo=TRUE, eval=params$show_answers, include=params$show_answers}
civic.turnout <- mean(subset(social, treat_civic == 1)$outcome_voted)
hawthorne.turnout <- mean(subset(social, treat_hawthorne == 1)$outcome_voted)
self.turnout <- mean(subset(social, treat_self == 1)$outcome_voted)
neighbors.turnout <- mean(subset(social, treat_neighbors == 1)$outcome_voted)

c(civic.turnout, hawthorne.turnout, self.turnout, neighbors.turnout)
```



## Practice Coding Exercise 2: Random Assignment

In the in-class activity we checked if age differed between the Neighbors and Control groups.

Does voter-history (`g2000`, `g2002`, `p2004`) differ by treatment assignment?

\small That is, what is the turnout rate for each election across the 4 treatments and control group?

What does this tell us about random assignment?


## Practice Coding Exercise 2: Answers {.fragile}

Remember how we subset the data before:

```{r answer-q2, echo=TRUE, include=params$show_answers}
civic.data <- subset(social, treat_civic == 1)
hawthorne.data <- subset(social, treat_hawthorne == 1)
self.data <- subset(social, treat_self == 1)
neighbors.data <- subset(social, treat_neighbors == 1)
```

\textcolor{blue}{Why did we need to do this for this problem as well?}

\pause

Because we wanted to know treatment assignment across several groups.


## g2000 Random Assignment {.fragile}

:::::::::::::: {.columns}
::: {.column width="50%"}
```{r, echo=TRUE, include=params$show_answers}
table(civic.data$g2000)
table(hawthorne.data$g2000)
```
:::
::: {.column width="50%"}
```{r, echo=TRUE, include=params$show_answers}
table(self.data$g2000)
table(neighbors.data$g2000)
```
:::
::::::::::::::

\vspace{1.5em}

\textcolor{blue}{What does this mean?}

\pause

There are similar values across each condition which means random assignment of treatment likely worked

## g2002 Random Assignment {.fragile}

:::::::::::::: {.columns}
::: {.column width="50%"}
```{r, echo=TRUE, include=params$show_answers}
table(civic.data$g2002)
table(hawthorne.data$g2002)
```
:::
::: {.column width="50%"}
```{r, echo=TRUE, include=params$show_answers}
table(self.data$g2002)
table(neighbors.data$g2002)
```
:::
::::::::::::::

## p2004 Random Assignment {.fragile}

:::::::::::::: {.columns}
::: {.column width="50%"}
```{r, echo=TRUE, include=params$show_answers}
table(civic.data$p2004)
table(hawthorne.data$p2004)
```
:::
::: {.column width="50%"}
```{r, echo=TRUE, include=params$show_answers}
table(self.data$p2004)
table(neighbors.data$p2004)
```
:::
::::::::::::::


## Acknowledgements

- Special thanks to Will Halm and prior GSIs for providing materials off of which some of these slides are based.
